---
title: "ESM 206 Lab 5"
author: "rubyhg"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# New markdown

## Objectives:

- Getting counts for different groups
- Use {lubridate} to parse dates
- Find confidence intervals & do t-tests with t.test()
- Heatmap with geom_tile()

# first thing is to attach packages at the top of the new markdown

```{r, include = FALSE}

library(tidyverse)
library(here)
library(janitor)
```

```{r}
# Import lobster data!

lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"), na = "-99999") %>%
  clean_names()
  

```

Use the tidyr::uncount() function to expand a frequency variable
```{r}
lobster_tidy <- lobster_abundance %>%
  tidyr::uncount(lobster_count)
```


Yay! Now each lobster has its own row. This is a tidy format.

Let's look at this a bunch of ways:
```{r}

ggplot(data = lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site),
              width = 0.2,
              alpha = 0.3)

```

```{r}
ggplot(data = lobster_tidy, aes(x = site, y = size_mm)) +
  geom_violin(aes(color = site),
              width = 0.2,
              alpha = 0.3)

```

```{r}

ggplot(data = lobster_tidy, aes(x = site, y = size_mm)) +
  geom_violin(aes(color = site),
              alpha = 0.3)

```

```{r}

ggplot(data = lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scales = "free")

```

```{r}

ggplot(data = lobster_tidy, aes(sample = size_mm)) +
  geom_qq(size = 0.5) +
  facet_wrap(~site, scales = "free")

```



Use "lubridate" to parse dates and times:

First, need to make "date" a variable not a character
> class(lobster_tidy$date)
[1] "character"
> lobster_date <- lobster_tidy %>%
+   mutate{
I'm going to add a new column with 'mutate()', that contains my date as an actual date

```{r}
lobster_date <- lobster_tidy %>%
  mutate(
    date_new = lubridate::mdy(date)
  )
     
```

> View(lobster_date)
> class(lobster_date$date_new)
[1] "Date"


Parse "date" column to get the differenct pieces (month, year) separated
```{r}
lobster_parse_date <- lobster_date %>%
  mutate(
    obs_month = lubridate::month(date_new, label = TRUE),
    obs_year = lubridate::year(date_new)
  )
```

label = TRUE to see month name (below shows that months are in order, this is how you view that)
> class(lobster_parse_date$obs_month)
[1] "ordered" "factor" 
> levels(lobster_parse_date$obs_month)
 [1] "Jan" "Feb" "Mar" "Apr" "May" "Jun" "Jul" "Aug" "Sep" "Oct" "Nov" "Dec"

obs_year = lubridate::year(date_new) creates a new column for year

Count lobsters by different groupings... count function is great for categorical data if you want to get total numbers

?count (this format gives you a description of the term)

Let's say I want to count the number of lobsters by year and month

```{r}
lobster_ym <- lobster_parse_date %>%
  count(obs_year, obs_month)

lobster_ym

```

What about for each year and site?
```{r}
lobster_ym <- lobster_parse_date %>%
  count(obs_year, obs_month)

lobster_ysite <- lobster_parse_date %>%
  count(obs_year, site)

lobster_ysite

```
```{r}
lobster_ym <- lobster_parse_date %>%
  count(obs_year, obs_month)

lobster_ysite <- lobster_parse_date %>%
  count(obs_year, site)

lobster_ysite

lobster_site <- lobster_parse_date %>%
  count(site)

lobster_site

```

Count is only so good, what if you want summaries and means and observations? Use group_by

```{r}
lobster_ym <- lobster_parse_date %>%
  count(obs_year, obs_month)

lobster_ysite <- lobster_parse_date %>%
  count(obs_year, site)

lobster_ysite

lobster_site <- lobster_parse_date %>%
  count(site)

lobster_site

# group_by + summarize + n() (this makes a new chart!)
lobster_summary <- lobster_parse_date %>%
  group_by(site) %>%
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    sample_n = n()
  )

```

Confidence intervals and t-tests

Use the "t-test()" function to find confidence intervals and perform t-tests

We want to find the confidence interval for lobster sizes at Isla Vista beach

```{r}

ivee_lobster <- lobster_tidy %>%
  filter(site == "IVEE") %>%
  pull(size_mm)

# pull() shows specific values (just a list)

t.test(ivee_lobster)

#t.test() is a one sample t-test
# confidence int tells us something about the sampling distribution - where would we expect to samples to fall 95% of the time?
# sample mean (even if you took a ton more samples), we expect the sample to fall within the confidence interval 95% of the time
# not the probability that the true population will exist in this range, but that I would expect any sample I take to fall within this interval 95% of the time

```

Two-sample t-test:
Is there a significant difference in mean lobster lengths between Naples Reef and Mohawk Reef?

```{r}

napl_sample <- lobster_tidy %>%
  filter(site == "NAPL") %>%
  pull(size_mm)


mohk_sample <- lobster_tidy %>%
  filter(site == "MOHK") %>%
  pull(size_mm)

# to use a two-way t-test, include both variables in t.test()
mohk_napl_ttest <- t.test(napl_sample, mohk_sample)
mohk_napl_ttest

# p-value = almost at zero. there is only a x probability that I could have taken two samples from pops with the same mean, and found by random chance with means of that difference...?

```

Mean lobster size differed significantly between Mohawk and Naples reefs (t(`r mohk_napl_ttest$parameter`) = 19.849)

You don't just want to type the degrees of freedom, you could make an error - you want to use "inline referencing," which is the blue above

Maybe an easier way to run a two sample t-test...
```{r}
lobster_2sample <- lobster_tidy %>%
  filter(site %in% c("NAPL","MOHK"))


ttest_2 <- t.test(size_mm ~ site, data = lobster_2sample)
ttest_2
```

Make a geom_tile heatmap

```{r}
ggplot(data = lobster_ysite, aes(x = obs_year, y = site)) +
  geom_tile(aes(fill = n))
```
